Particle Swarm Optimization for Crystallography

### 2009-01-07 crystallography/paper.tex
Some initial work on an intro
Reworked introduction, started on results

\begin{abstract}
TODO.
\end{abstract}

@@

TODO.
Particle swarm optimization is most often used with functions that are quickly
evaluated.  Many real-life problems take much longer.  What does one do when 
presented with a function that takes several minutes or hours to evaluate?

In this paper we present a crystallography problem whose function evaluation
takes several minutes to evaluate.  We discuss issues in solving this problem
specifically, and in optimizing expensive functions in general with particle 
swarm optimization.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Crystallography

\section{Crystallography}
\label{sec:crystallography}

FIXME.
TODO: Where to put the butterfly problem definition? Should I describe 
simulated annealing or gradient descent, at least briefly?

The point of crystallography is to determine the structure of a crystal.  This
is done by shining x-rays through the crystal and observing the patterns that
come out.  Those patterns are then compared to the output of a model of the 
crystal -- if the two match, the model can be considered correct.  The model
may consist of the positions of all of the atoms in the crystal, the 
distortion coefficients of the materials, the length and width of nanopolar
domains, or various other parameters.  The task of finding the parameters to 
the model that best fit the data is well suited to optimization techniques.  
The function to be optimized is an error function -- the average difference
between the model output and the data, for every data point.

There are commercial packages that readily solve some crystallography problems,
such as (ask Dr. Campbell), that uses simulated annealing for specific types 
of problems.  Gradient descent is also a commonly used method for solving these
problems.  Particle swarm optimization, however (to our knowledge?), has not 
been used.

We encountered a problem for which gradient descent proved uneffective.

To solve crystallography problems, one must fit the parameters of the model
to the data.  The more data is used, the more accurate the result will be, but
with large amounts of data the evaluation time can be inhibitively large.  In
our first attempts at solving this problem, the function evaluation took close
to an hour.  We then trimmed the amount of data we were using, and the time for
a function evaluation decreased to about four minutes.

@@

@@

\bibliographystyle{plain}
\bibliography{%
    \bib{mapreduce/dean-osdi04},%
    \bib{pso/clerc-tec02},%
    \bib{pso/kennedy-icnn95},%
    \bib{pso/bratton-sis07},%
    \bib{aml/mcnabb-cec07}}

>>>>

\begin{abstract}
FIXME.

In this paper we present a crystallography problem whose function evaluation
takes several minutes to evaluate.  We discuss issues in solving this problem
specifically, and in optimizing expensive functions in general with particle 
swarm optimization.
\end{abstract}

@@

FIXME.

Crystallography is the science of determining the structure of a crystal.  This
is done by shining x-rays through the crystal and observing the patterns that
come out, generally consisting of intensities of light at various points in
space.  Those patterns are then compared to the output of a model of the
crystal -- the predicted intensities of light at the same points in space.  If
the two match, the model can be considered correct.  The model may consist of
the positions of all of the atoms in the crystal, the distortion coefficients
of the material, the length and width of nanopolar domains, or various other
parameters.  The task of finding the parameters to the model that best fit the
data is well suited to optimization techniques.  The function to be optimized
is an error function -- the average difference between the model output and the
data, for every data point.

There are commercial packages that readily solve some crystallography problems,
such as (ask Dr. Campbell), that uses simulated annealing for specific types of
problems.  Gradient descent is also a commonly used method for solving these
problems.  Particle swarm optimization (PSO), described in a later section, has
also been used, but only in limited contexts \cite{liu-anc05}.

To solve crystallography problems, one must fit the parameters of the model to
the data.  The more data is used, the more accurate the result will be, but
with large amounts of data the error function evaluation time can be
inhibitively large, easily reaching minutes or even hours.  In such cases,
performing any kind of optimization on a single machine is infeasible; parallel
techniques are needed.

We present a crystallography problem whose error function evaluation takes
several minutes. We then present our solution to the problem, using MRPSO, a
parallel particle swarm optimization implementation \cite{mcnabb-cec07}.  The
issues we encountered and solved dealing with this crystallography problem are
applicable to any problem whose function evaluation takes large amounts of
time.

Section~\ref{sec:butterfly} introduces the crystallography problem we solved.
Section~\ref{sec:pso} explains the particle swarm optimization algorithm, and
section~\ref{sec:mapreduce} explains MapReduce and PSO in the context of
MapReduce.  Section~\ref{sec:results} then describes our solution to the
problem.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Butterfly Problem
\section{Crystallography Problem}
\label{sec:butterfly}
TODO.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Results
\section{Results (Maybe Issues with Optimization?)}
\label{sec:results}
TODO.

In the course of solving this problem, it was discovered that a few of the 
variables were correlated.  In initial attempts at optimization, the algorithm
never converged -- the function value reached a certain point and stayed close
to that point while a couple of the parameters fluctuated greatly in a 
correlated fashion.  Figure (make figure) shows a range of values that two of
parameters took on, without changing the function value more than (check number)
.5, or this \%.  Upon noticing this empirical correlation, the function was
more closely examined, and a theoretical correlation was also discovered.  
Given that the parameters were correlated and the function had too many free
variables, one of the parameters was fixed at a reasonable value, and the 
optimization algorithm successfully and quickly solved the problem.

Work has been done on the use of particle swarm optimization in parallel with 
large numbers of particles \cite{mcnabb-gecco09}.

@@

\bibliographystyle{plain}
\bibliography{%
    \bib{mapreduce/dean-osdi04},%
    \bib{pso/clerc-tec02},%
    \bib{pso/kennedy-icnn95},%
    \bib{pso/bratton-sis07},%
    \bib{aml/mcnabb-cec07},%
    \bib{pso/crystallography/liu-anc05}}
